Preventing a Nuclear ISIL: \\Recommendations for United States Policy

### 2015-05-11 draft.tex
Updated most of the list of acronyms. Pushed this commit to Red Team for final edits.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

ISIL's principal source of cash flow comes from the control and sale of oil, which brings in about \$1 million a day. ISIL is also financed through kidnappings, extortion, criminal activities, and a small amount from external donations \cite{Lister2014}. ISIL employs roughly 30,000 fighters, including 19,000 foreign fighters from 90 different countries, paying them roughly \$350-\$500 a month. This comes out to \$10 million per month. ISIL also recruits fighters in the region by providing electricity and other services to local residents \cite{TheEditorialBoard2015}. A new report by Financial Action Task Force (FATF) finds that it is unclear whether ISIL's revenue collection will be sustainable over time, and that ISIL will have to gain more territory to remain financially viable \cite{TheEditorialBoard2015,Report2015}.

@@

\begin{itemize}
 \item Use the architecture of the Expanded Threat Reduction Initiative (ETR) of the Clinton Administration to fund employment and research for nuclear scientists through the State Department. 
 \item Continue to combat ISIL propaganda with media outreach, in order to lessen support for the group that may spread to nuclear scientists.
\end{itemize}

In order to reduce monetary incentives for scientists to join ISIL, the most effective measure is to ensure that nuclear scientists have steady employment and research opportunities. A Congressional Research Report for Congress from 1999 (obtained by WikiLeaks) spells out the Clinton administration's approach to this problem by evaluating the administration's budget request from \$4.5 billion over 5 years for the Expanded Threat Reduction Initiative (ETR) \cite{Woolf1999}. In the report, Amy Woolf and Curt Tarnoff described the steps proposed in the budget request:

@@

\ldots The State Department manages several programs that offer NIS scientists employment and research opportunities, so that they will be less likely to sell their knowledge to rogue nations. The International Science and Technology Center (ISTC) in Moscow and the Science and Technology Center in Ukraine (STCU), multilateral programs supported by the EU, Japan, and others, fund research in such areas as nuclear reactor safety, medicine, civil aviation, and energy production. A \enquote{Partners Program} links NIS scientists with contributing U.S. private sector companies, such as General Electric and Lockheed Martin, and academic institutes to assist their own R\&D efforts. The Civilian Research and Development Foundation (CRDF) is a private, non-profit foundation, originally funded by matching grants from DoD and George Soros. It supports U.S. collaboration and exchanges with former NIS defense scientists on civilian basic and applied research, and supports projects with commercial applications. The State Department also implements a pilot project for biological weapons scientists to \enquote{redirect} their expertise toward commercial endeavors. Among other activities, it has established a collaborative program between the Department of Agriculture's Agricultural Research Service and Russian scientists on agriculture research, and between the Department of Health and Human Services' Centers for Disease Control, National Institutes of Health, and Food and Drug Administration and Russian biomedical scientists to conduct research on infectious diseases. }

@@

Existing Policy: \textcolor{red}{Second Line of Defense (SLD) Core Program}

@@

Existing Policy:  \textcolor{red}{IAEA Monitoring \& the Additional Protocol (AP)}

@@

The lead U.S. agency responsible for implementing this policy will be the U.S. Department of State (DoS) supported by the U.S. Defense Threat Reduction Agency (DTRA). Preventing ISIL from acquiring a nuclear weapon, or at the very least SNM, falls under the purview of both the Bureau of Counterterrorism (CT) and the Bureau of International Security and Nonproliferation (ISN) within the DoS. Under this policy strategy, ISN is the principal actor in the DoS as they are the primary U.S. interface with the IAEA on the NPT \cite{Circular1970}. DTRA is a Combat Support Agency that supports the Combatant Commanders and Services in responding to any WMD threat. One aspect of this support is managing a research and development portfolio to develop tools and capabilities to detect and respond to WMD threats. DTRA is the agency within the U.S. government that works to verify treaties related to WMD are being carried out by all signatories, and thus will be able to support the DoS ISN in this policy \cite{DefenseThreatReductionAgency}.

@@

The time for the implementation of an  expert group specializing in nuclear training can range from six months to a year, based on past experience.  Training groups can be built from the same groups in the IAEA that hold nuclear security and education events. The U.S Department of Transportation and NNSA (National Nuclear Security Administration) will work in conjunction with countries to help them implement safety across borders to prevent the movement (theft) of nuclear material.

@@

\begin{figure}
 \centering
 \includegraphics[trim = 0cm 0.1cm 0cm 0cm, clip,scale=0.4]{./figures/shield_time.png}
   \caption{Time needed to reach a gamma count of highly enriched uranium (HEU) 5 standard deviations above the background for a given lead shield thickness \cite{Glaser2007}.}
     \label{fig:shield_time}
\end{figure}

@@

The UN is the largest international organization that can take actions on the issues of security at a global level. Several divisions in the UN address nuclear security issues including the General Assembly (GA), the Security Council (SC), and the Department of Disarmament Affairs (DDA). The GA can make recommendations to Member States and the SC about situations that are likely to endanger international peace and security. The SC consists of five permanent member states (China, France, Russia, United Kingdom, and the United States) and 10 non-permanent members. The Security Council is responsible for formulating plans for the regulation of armaments. The DDA provides substantive and organizational support to UN Member States for disarmament norm-setting through the work of the GA and other bodies. The DDA's Weapons of Mass Destruction (WMD) Branch supports multilateral efforts to strengthen the international norm on disarmament and non-proliferation of WMD and cooperates with other UN agencies, such as the International Atomic Energy Agency (IAEA).

>>>>

AP - Additional Protocol

@@

DoE - Department of Energy

DoJ - Department of Justice

@@

DoT - Department of Treasury

DTRA - Defense Threat Reduction Agency

ETRI - Expanded Threat Reduction Initiative

FATF - Financial Action Task Force

@@

GTRI - Global Threat Reduction Initiative

HEU - Highly Enriched Uranium

@@

IS - Islamic State

@@

ISN - Bureau of International Security and Nonproliferation

@@

ISTC - International Science and Technology Center

@@

NNSA - National Nuclear Security Administration

@@

NTNFC - National Technical Nuclear Forensics Center

@@

OMB - Office of Management and Budget

@@

SLD - Second Line of Defense

@@

ISIL's principal source of cash flow comes from the control and sale of oil, which brings in about \$1 million a day. ISIL is also financed through kidnappings, extortion, criminal activities, and a small amount from external donations \cite{Lister2014}. ISIL employs roughly 30,000 fighters, including 19,000 foreign fighters from 90 different countries, paying them roughly \$350-\$500 a month. This comes out to \$10 million per month. ISIL also recruits fighters in the region by providing electricity and other services to local residents \cite{TheEditorialBoard2015}. A new report by the  Financial Action Task Force (FATF) finds that it is unclear whether ISIL's revenue collection will be sustainable over time, and that ISIL will have to gain more territory to remain financially viable \cite{TheEditorialBoard2015,Report2015}.

@@

\begin{itemize}
 \item Use the architecture of the Expanded Threat Reduction Initiative (ETRI) of the Clinton Administration to fund employment and research for nuclear scientists through the State Department. 
 \item Continue to combat ISIL propaganda with media outreach, in order to lessen support for the group that may spread to nuclear scientists.
\end{itemize}

In order to reduce monetary incentives for scientists to join ISIL, the most effective measure is to ensure that nuclear scientists have steady employment and research opportunities. A Congressional Research Report for Congress from 1999 (obtained by WikiLeaks) spells out the Clinton administration's approach to this problem by evaluating the administration's budget request from \$4.5 billion over 5 years for the Expanded Threat Reduction Initiative  \cite{Woolf1999}. In the report, Amy Woolf and Curt Tarnoff described the steps proposed in the budget request:

@@

\ldots The State Department manages several programs that offer NIS scientists employment and research opportunities, so that they will be less likely to sell their knowledge to rogue nations. The International Science and Technology Center (ISTC) in Moscow and the Science and Technology Center in Ukraine (STCU), multilateral programs supported by the EU, Japan, and others, fund research in such areas as nuclear reactor safety, medicine, civil aviation, and energy production. A \enquote{Partners Program} links NIS scientists with contributing U.S. private sector companies, such as General Electric and Lockheed Martin, and academic institutes to assist their own R\&D efforts. The Civilian Research and Development Foundation  is a private, non-profit foundation, originally funded by matching grants from DoD and George Soros. It supports U.S. collaboration and exchanges with former NIS defense scientists on civilian basic and applied research, and supports projects with commercial applications. The State Department also implements a pilot project for biological weapons scientists to \enquote{redirect} their expertise toward commercial endeavors. Among other activities, it has established a collaborative program between the Department of Agriculture's Agricultural Research Service and Russian scientists on agriculture research, and between the Department of Health and Human Services' Centers for Disease Control, National Institutes of Health, and Food and Drug Administration and Russian biomedical scientists to conduct research on infectious diseases. }

@@

Existing Policy: \textcolor{red}{Second Line of Defense Core Program}

@@

Existing Policy:  \textcolor{red}{IAEA Monitoring \& the Additional Protocol}

@@

The lead U.S. agency responsible for implementing this policy will be the U.S. Department of State  supported by the U.S. Defense Threat Reduction Agency. Preventing ISIL from acquiring a nuclear weapon, or at the very least SNM, falls under the purview of both the Bureau of Counterterrorism and the Bureau of International Security and Nonproliferation (ISN) within the DoS. Under this policy strategy, ISN is the principal actor in the DoS as they are the primary U.S. interface with the IAEA on the NPT \cite{Circular1970}. DTRA is a Combat Support Agency that supports the Combatant Commanders and Services in responding to any WMD threat. One aspect of this support is managing a research and development portfolio to develop tools and capabilities to detect and respond to WMD threats. DTRA is the agency within the U.S. government that works to verify treaties related to WMD are being carried out by all signatories, and thus will be able to support the DoS ISN in this policy \cite{DefenseThreatReductionAgency}.

@@

The time for the implementation of an  expert group specializing in nuclear training can range from six months to a year, based on past experience.  Training groups can be built from the same groups in the IAEA that hold nuclear security and education events. The U.S Department of Transportation and NNSA  will work in conjunction with countries to help them implement safety across borders to prevent the movement (theft) of nuclear material.

@@

\begin{figure}
 \centering
 \includegraphics[trim = 0cm 0.1cm 0cm 0cm, clip,scale=0.4]{./figures/shield_time.png}
   \caption{Time needed to reach a gamma count of highly enriched uranium  5 standard deviations above the background for a given lead shield thickness \cite{Glaser2007}.}
     \label{fig:shield_time}
\end{figure}

@@

The UN is the largest international organization that can take actions on the issues of security at a global level. Several divisions in the UN address nuclear security issues including the General Assembly (GA), the Security Council (SC), and the Department of Disarmament Affairs (DDA). The GA can make recommendations to Member States and the SC about situations that are likely to endanger international peace and security. The SC consists of five permanent member states (China, France, Russia, United Kingdom, and the United States) and 10 non-permanent members. The Security Council is responsible for formulating plans for the regulation of armaments. The DDA provides substantive and organizational support to UN Member States for disarmament norm-setting through the work of the GA and other bodies. The DDA's Weapons of Mass Destruction (WMD) Branch supports multilateral efforts to strengthen the international norm on disarmament and non-proliferation of WMD and cooperates with other UN agencies, such as the IAEA.
